Nitric oxide feedback to ciliary photoreceptor cells orchestrates UV avoidance

### 2023-05-23 Manuscript/Manuscript.Rmd
MS: rewrote introdution, added refs

---
title: Nitric-oxide-feedback to ciliary photoreceptor cells orchestrates UV avoidance
  in zooplankton
author: 
  - name: "Kei Jokura<sup>1,`*`</sup>, Nobuo Ueda<sup>2</sup>, Martin Gühmann<sup>3</sup>, Luis Alfonso Yañez-Guerra<sup>1</sup>, Piotr Słowiński<sup>1</sup>, Kyle C. A. Wedgewood<sup>1</sup> and Gáspár Jékely<sup>1,4,`*`</sup>"
    affiliation: "<sup>1</sup>Living Systems Institute, University of Exeter, Stocker road, Exeter, EX4 4QD, United Kingdom <sup>2</sup>Okinawa Institute of Science and Technology, Okinawa, Japan, <sup>3</sup>School of Biological Sciences, University of Bristol, Bristol, United Kingdom, <sup>4</sup>Centre for Organismal Studies (COS), University of Heidelberg, 69120 Heidelberg, Germany  <br> <sup>*</sup>Correspondence`:` k.jokura @ exeter.ac.uk, gaspar.jekely @ cos.uni-heidelberg.de"
output:
  html_document:
    theme: lumen
  pdf_document:
    latex_engine: lualatex
    template: default
  word_document: default
geometry: left=0.8in,right=0.8in,top=0.8in,bottom=0.5in
mainfont: Arial
monofont: Times New Roman
documentclass: article
classoption: onecolumn
linestretch: 1.2
fontsize: 10pt
bibliography: references.bib
csl: elife.csl
editor_options: 
  markdown: 
    wrap: sentence
---

@@

Nitric oxide (NO) generated by nitric-oxide synthase (NOS) is a key regulator of animal physiology.
Here we uncover a role for NO in shaping circuit dynamics to orchestrate light-avoidance behaviour.
We studied UV-avoidance mediated by brain ciliary photoreceptors (cPRCs) in larval *Platynereis dumerilii*, a marine annelid.
We found NOS specifically expressed in interneurons (INNOS) postsynaptic to cPRCs.
Stimulation of cPRCs by violet light leads to cPRC inhibition but concomitant INNOS activation and NO production.
NO feeds back to cPRCs and triggers their delayed activation through an unconventional guanylyl cyclase.
This results in the activation of projection interneurons and the inhibition of serotonergic ciliomotor neurons.
In *NOS* mutants, NO feedback and projection-neuron activation do not occur and avoidance behaviour is defective.
By mathematical modelling, we could capture the circuit dynamics in both wild-type and mutant larvae.
Our results reveal how NO-mediated neuroendocrine signalling gates a synaptic circuit to initiate light-avoidance behaviour.

@@

Neuronally produced nitric oxide (NO) can act as a diffusible neuromodulator.
This small free-radical molecule is synthesized by nitric oxide synthase (NOS).
The most major target of NO signalling in the neuronal circuit is soluble guanylate cyclase (sGC).
sGC activated by NO produces cGMP, which acts as a second messenger and regulates neuronal activity.
In the mammalian retina, NO is an essential signalling molecule and one of the most well-studied [@Cudeiro1999].
NO produced in amacrine cells, one of the retinal cells[@jacoby2018], activates sGCs (general targets of NO) in pyramidal bipolar cells, the presynaptic nerves, via retrograde signalling; NO-dependent increases in cGMP have been reported to regulate neurotransmission [@Vielma2014][@Wei2012].
It has also been shown that in Drosophila, object features are differentially stored and that visual orientation memory for objects out of view is processed in the central complex of the fly brain, with ring neurons in the oval body involved in memory formation.
Second messengers in the ellipsoid are required for directional memory and nitric oxide (NO) has been implicated in memory formation.
cGMP-dependent signalling pathways via sGC cause activation of transcription factors, which are necessary for the ability to build orientation memories[@Kuntz2017].
NO as a neuromodulator has been reported to be involved in the regulation of vision and memory via the NO-sGC-cGMP pathway in neural circuits, but the role of NO signalling in the regulation of neural circuit dynamics is not well understood to the molecular level mechanisms.

To investigate NO signalling in the modulation of neural circuit dynamics, we studied the larval stages of the marine annelid model *Platynereis dumerilii* [@ozpolat2021].
We uncovered an essential function for NO in these larvae in UV/violet-light avoidance.
In *Platynereis* larvae, UV/violet avoidance is characterised by downward swimming and is mediated by the UV/violet-responding brain ciliary photoreceptor cells (cPRCs).

The cPRCs express a vertebrate-type ciliary opsin, c-opsin1 [@Arendt2004].
This opsin forms a UV-absorbing bistable photopigment with an absorption maximum around 384 nm [@tsukamoto2017, @Veraszto2018].
*c-opsin1* knockout larvae show defective UV/violet avoidance [@Veraszto2018].
Here we show that *Platynereis* *NOS* is expressed in interneurons of the cPRC circuitry and is required for UV/violet-avoidance behaviour.
By combining calcium imaging across the fully-mapped cPRC circuit, genetic perturbations and mathematical modelling, we describe how NO tunes circuit dynamics through non-synaptic reverse signalling to cPRCs.
This delayed neuroendocrine feedback is essential for circuit output, and integrates UV/violet input and also induces a short-term memory manifest in altered circuit activity.

#In adult worms, UV light and c-opsin1 modulate locomotion and maturation [@Veedin2021].

@@

Next, we wanted to directly test whether NO signalling is involved in UV/violet avoidance.
To achieve this, we generated two *Platynereis* *NOS* knockout lines with the CRISPR/Cas9 method.
We recovered two deletions (NOSΔ11/Δ11 and Δ23/Δ23), both leading to frame-shift and an early stop codon and thus likely representing null alleles **(Figure 3---figure supplement 1A)**.
We could establish a homozygous line for both mutations indicating that NOS is not an essential gene.
To quantify UV avoidance, we recorded the trajectories of freely swimming wild type and mutant larvae in vertical columns, illuminated laterally from two opposite sides with 395 nm UV light **(Figure 3A and Figure 3---figure supplement 1B)**.
As previously shown, wild-type larvae swim downward following non-directional UV/violet light stimulation [@Veraszto2018].
In contrast, both two- and three-day-old homozygous *NOS*-mutant larvae showed a strongly diminished UV-avoidance response **(Figure 3A, B and Figure 3---figure supplement 1B,C)**.
This phenotype is similar to the defective UV-avoidance of *c-opsin1* mutant larvae and reveals a requirement for NOS in UV-avoidance behaviour [@Veraszto2018].
Wild type but not mutant larvae also showed an increase in swimming speed under UV light that may be due to downward swimming trajectories (swimming with gravity) **(Figure 3B and Figure 3---figure supplement 1C)**.
We also tested directional phototaxis, by exposing larvae to 480 nm directional and collimated light from the top of the column.
Three-day-old but not two-day-old *NOS*-mutant larvae also showed reduced phototactic behaviour, suggesting a function for *NOS* in the visual eyes that mediate three-day-old phototaxis [@Randel2014a] **(Figure 3D and Figure 3---figure supplement 1G)**.
To distinguish between an acute or developmental function of NOS in light responses, we next tested larvae exposed to the NOS inhibitor L-NAME.
Larvae incubated for 5 min in 0.1 mM or 1 mM L-NAME showed a dose-dependent inhibition of UV avoidance.
In contrast, phototaxis was not effected **(Figure 3C, E)**.
Overall, our results indicate an acute requirement for NOS signalling in UV-avoidance and a possible indirect, developmental role in visual phototaxis.

>>>>

---
title: Nitric oxide feedback to ciliary photoreceptor cells orchestrates UV avoidance
  in zooplankton
author: 
  - name: "Kei Jokura<sup>1,`*`</sup>, Nobuo Ueda<sup>2</sup>, Martin Gühmann<sup>3</sup>, Luis Alfonso Yañez-Guerra<sup>1</sup>, Piotr Słowiński<sup>1</sup>, Kyle C. A. Wedgewood<sup>1</sup> and Gáspár Jékely<sup>1,4,`*`</sup>"
    affiliation: "<sup>1</sup>Living Systems Institute, University of Exeter, Stocker road, Exeter, EX4 4QD, United Kingdom <sup>2</sup>Okinawa Institute of Science and Technology, Okinawa, Japan, <sup>3</sup>School of Biological Sciences, University of Bristol, Bristol, United Kingdom, <sup>4</sup>Centre for Organismal Studies (COS), University of Heidelberg, 69120 Heidelberg, Germany  <br> <sup>*</sup>Correspondence`:` k.jokura @ exeter.ac.uk, gaspar.jekely @ cos.uni-heidelberg.de"
output:
  html_document:
    theme: lumen
  pdf_document:
    latex_engine: lualatex
    template: default
  word_document: default
geometry: left=0.8in,right=0.8in,top=0.8in,bottom=0.5in
mainfont: Arial
monofont: Times New Roman
documentclass: article
classoption: onecolumn
linestretch: 1.2
fontsize: 10pt
bibliography: references.bib
csl: elife.csl
editor_options: 
  markdown: 
    wrap: sentence
---

@@

Nitric oxide (NO) generated by nitric-oxide synthase (NOS) is a key regulator of animal physiology.
Here we uncover a role for NO in shaping circuit dynamics to orchestrate light-avoidance behaviour.
We studied UV/violet light avoidance mediated by brain ciliary photoreceptors (cPRCs) in larval *Platynereis dumerilii*, a marine annelid.
We found NOS expressed in interneurons (INNOS) postsynaptic to cPRCs.
Stimulation of cPRCs by violet light leads to cPRC inhibition but concomitant INNOS activation and NO production.
NO feeds back to cPRCs and triggers their delayed activation through an unconventional guanylyl cyclase.
This results in the activation of projection interneurons and the inhibition of serotonergic ciliomotor neurons.
In *NOS* mutants, NO feedback and projection-neuron activation do not occur and avoidance behaviour is defective.
By mathematical modelling, we could recapitulate phototransduction and circuit dynamics in both wild-type and mutant larvae.
Our results reveal how NO-mediated retrograde signalling gates a synaptic circuit to initiate light-avoidance behaviour.

@@

In nervous systems, synaptic transmission and volume transmission together shape circuit dynamics.
While synaptic transmission occurs at specialised contact sites, volume transmission is characterised by the delocalised release of diverse diffusive neuromodulators.

Nitric oxide (NO) is one such modulator with unique physical and signalling properties.
This free radical synthesized by nitric oxide synthase (NOS) is short-lived and can diffuse across biological membranes [@Cudeiro1999; @thomas2015].
Canonical NO signalling involves the activation of neuronal NOS following calcium increase, NO production and diffusion, and the NO-dependent activation of soluble guanylate cyclases (sGC) in target cells leading to cGMP production [@Bredt1990; @Hoelscher1997].
Given that NOS activation is calcium dependent and NOS shows neuron-type-specific expression [@Gibbs1998a; @Wildemann1999; @Mobley2022], NO action can lead to the activity-dependent modulation of neural circuits at specific sites [@Wang2007; @Vielma2014; @jacoby2018] .

NO signalling has been extensively studied in the vertebrate retina where NOS is expressed in amacrine, ganglion as well as other cells [@Cudeiro1999; @Wang2007; @jacoby2018].
The actions of NO can be diverse and not always involve the canonical sGC-cGMP pathway [@Tooker2013; @Wei2012; @jacoby2018].
Defective NO signalling in NOS knockouts for example leads to a decreased sensitivity of retinal ganglion cells to light stimulation [@Wang2007].
However, due to the complex expression of NOS in vertebrates and the diversity of its functions, it has been challenging to link the neurophysiological effects of NO to behaviour change.

In the whole organism context, NO signalling has often been studied in diverse marine invertebrates, where NO can regulate larval settlement and metamorphosis [@leise2001a; @ueda2016; @Locascio2022; @song2021; @Zhang2012].
While NOS-expressing neurons potentially responsible for these effects have been reported [@bishop2007; @Locascio2022], it has not been possible to link these NO-dependent effects on behaviour or life-cycle transitions to neuronal activity and function.

Due to these difficulties, we still know little at the whole organism level about how NO production relates to stimulus conditions, how it shapes circuit activity at specific neuron types, and how NO-dependent modulation relates to behaviour.

To investigate NO function in neural circuit dynamics and behaviour, here we study larvae of the marine annelid model *Platynereis dumerilii* [@ozpolat2021].
*Platynereis* has emerged as a systems neuroscience model where behaviour and activity imaging can be combined with genetic manipulations and investigated in the context of a whole-body synaptic connectome and gene expression atlases [@Vergara2021; @Veraszto2017; @veraszto2020; @ozpolat2021].

Here we uncovered an essential function for NO signalling in larval UV/violet-light avoidance behaviour.
In *Platynereis* larvae, UV/violet avoidance is mediated by brain ciliary photoreceptor cells (cPRCs) and is characterised by downward swimming [@Veraszto2018].
The cPRCs express a ciliary-type opsin, c-opsin1 [@Arendt2004] that forms a UV-absorbing bistable photopigment with an absorption maximum around 384 nm [@tsukamoto2017, @Veraszto2018, @Veedin2021].
Upon UV/violet exposure, the cPRC show a characteristic biphasic calcium response that is *c-opsin1* dependent.
UV/violet avoidance behaviour is also *c-opsin1*-dependent and is defective in *c-opsin1* mutants [@Veraszto2018].
Here we show that *Platynereis* *NOS* is expressed in interneurons of the cPRC circuitry and is required for UV/violet-avoidance.
By combining calcium imaging across the fully-mapped cPRC circuit [@Veraszto2018] with genetic perturbations and mathematical modelling, we describe how NO tunes circuit dynamics through non-synaptic retrograde signalling to cPRCs.
This delayed neuroendocrine feedback integrates UV/violet exposure to induce a short-term memory manifested in altered circuit activity and an aversive behavioural response.

@@

Next, we wanted to directly test whether NO signalling is involved in UV/violet avoidance.
To achieve this, we generated two *Platynereis* *NOS* knockout lines with the CRISPR/Cas9 method.
We recovered two deletions (NOSΔ11/Δ11 and Δ23/Δ23), both leading to frame-shift and an early stop codon and thus likely representing null alleles **(Figure 3---figure supplement 1A)**.
We could establish a homozygous line for both mutations indicating that NOS is not an essential gene.
To quantify UV avoidance, we recorded the trajectories of freely swimming wild type and mutant larvae in vertical columns, illuminated laterally from two opposite sides with 395 nm UV light **(Figure 3A and Figure 3---figure supplement 1B)**.
As previously shown, wild-type larvae swim downward following non-directional UV/violet light stimulation [@Veraszto2018].
In contrast, both two- and three-day-old homozygous *NOS*-mutant larvae showed a strongly diminished UV-avoidance response **(Figure 3A, B and Figure 3---figure supplement 1B,C)**.
This phenotype is similar to the defective UV-avoidance of *c-opsin1* mutant larvae and reveals a requirement for NOS in UV-avoidance behaviour [@Veraszto2018].
Wild type but not mutant larvae also showed an increase in swimming speed under UV light that may be due to downward swimming trajectories (swimming with gravity) **(Figure 3B and Figure 3---figure supplement 1C)**.
We also tested directional phototaxis, by exposing larvae to 480 nm directional and collimated light from the top of the column.
Three-day-old but not two-day-old *NOS*-mutant larvae also showed reduced phototactic behaviour, suggesting a function for *NOS* in the visual eyes that mediate three-day-old phototaxis [@Randel2014a] **(Figure 3D and Figure 3---figure supplement 1G)**.
To distinguish between an acute or developmental function of NOS in light responses, we next tested larvae exposed to the NOS inhibitor L-NAME.
Larvae incubated for 5 min in 0.1 mM or 1 mM L-NAME showed a dose-dependent inhibition of UV avoidance.
In contrast, phototaxis was not effected **(Figure 3C, E)**.
Overall, our results indicate an acute requirement for NOS signalling in UV-avoidance and a possible indirect, developmental role in the visual system, reminiscent of the developmental role for NO signalling in the *Drosophila* eye [@Gibbs1998a].
